\section{Experimental Setup}

\paragraph{Data.}
The real pilot uses two public English GEC data sources. EXPECT is released under the MIT License; the versioned pilot file contains 300 real source/reference pairs and 320 source-reference ERRANT edits. JFLEG is released under CC-BY-NC-SA-4.0; we retain all four references and use reference 0 as the primary reference for this single-reference ERRANT pilot. Exact download commands, commits, and license notes are recorded in the project data statement.

\paragraph{Model predictions.}
The current pilot evaluates edits actually produced by public GEC models rather than treating reference corrections as model predictions. It includes GECToR RoBERTa base 5k as a sequence-to-edit system, T5 base grammar correction as a sequence-to-sequence system, and a small CoEdIT-large instruction-following editor branch. We store source, reference, prediction, model version, decoding configuration, and CPU runtime metadata for every prediction.

\paragraph{Edit extraction.}
We use ERRANT 3.0.2 with the small spaCy English model to extract structured edits from source/reference pairs and source/model-prediction pairs. Predicted edits are aligned against reference edits and classified as correct correction, wrong correction, or overcorrection; unmatched reference edits are stored separately as missed corrections. For transparency, we also compare ERRANT with the earlier token-diff extractor. Token diff exactly matches ERRANT edit spans and strings on 94.7\% of sampled sentences before error types are considered, confirming that ERRANT is necessary for linguistically typed edit boundaries.

\paragraph{Explanation pilot.}
The earlier source-reference template pilot is retained only as a leakage-control sanity check. For model-produced edits, EXPECT provides evidence-word and error-type labels but no natural-language explanation field, so we generate open-source explanation candidates with FLAN-T5-base from source, model prediction, and the predicted edit span. These candidates are not human gold. Explicit edit templates are retained only as an upper control for answer leakage.

\paragraph{Qwen3-8B method pilot.}
For the method line, we compare local open teachers under the same 80-edit setting. The teacher baselines are FLAN-T5-base, Qwen2.5-0.5B-Instruct, a Qwen2.5-1.5B probe, and Qwen3-8B with thinking disabled. Each full run produces two candidates per edit, one natural and one rule-grounded, for 160 candidates unless explicitly marked as a smaller probe. We then use Qwen3-8B for the RuleFaith pilot because it is the first local teacher to pass the conservative prefilter at non-trivial scale. The Qwen3 pilot records raw responses, parsed outputs, candidate IDs, prompt version, and repair provenance. We then run field-aware RuleFaith selection, target-masked validation, rule/evidence audit, deterministic targeted repair, and a final Codex/AI pseudo-validation pass. The pseudo-validation is used for internal triage only; it is not counted as human annotation.

\paragraph{Benchmark candidate.}
Round 07 constructs an automatic model-edit-level benchmark with 700 selected model-produced edits, 12.8k explanation/control instances, and 160 missed-edit diagnoses. The selected edits cover EXPECT and JFLEG, three model families, replace/insert/delete operations, and 41 ERRANT error types. These automatic labels support method development and leakage analysis; they do not replace human faithfulness annotation.

\paragraph{Baselines.}
The pilot runs random and majority controls, surface keyword matching, structured explicit extraction, reverse reconstruction, no-source reconstruction, target-masked reconstruction, leakage-adjusted reconstruction, TF-IDF similarity, an NLI lexical proxy, a rule/evidence verifier, counterfactual simulator baselines, and reranking baselines. Explicit templates and raw edit strings are reported only as leakage upper controls.

\paragraph{Metrics.}
For L1, we report Full Edit Exact Match, Span F1, Target Match, Operation Accuracy, Error Type Accuracy, and leakage-adjusted deltas. For L2, we report counterfactual behavior accuracy, Macro-F1, grouped bootstrap confidence intervals by source sentence, and preserve/change/cancel/competing-edit breakdowns. For L3, we report rule consistency, evidence span checks, contradiction rejection, and negative rejection rates. For reranking, we report pairwise ranking accuracy, MRR, top-1 automatic faithfulness, automatic grammatical-validity and helpfulness proxies, fluency proxy, explanation length, and reward-hacking indicators. All reported numbers are generated from JSON outputs under \texttt{results/}.
