\section{Introduction}

Grammatical error correction (GEC) systems are designed to transform learner or user-written text into a corrected version while preserving the intended meaning. Standard evaluations have historically emphasized whether the final corrected sentence matches reference corrections, while tools such as ERRANT make it possible to analyze edits and error types more directly. Recent work has expanded GEC beyond correction alone: systems may now generate explanations, highlight evidence, or provide natural-language feedback that is meant to help users understand why a correction was suggested.

This additional explanatory layer creates a different evaluation problem. A correction can be useful while its explanation is vague, misplaced, or inconsistent with the actual edit. For example, an explanation may mention subject-verb agreement even when the model changed an article, or it may describe replacing one word with another while the system actually inserted a missing preposition. Such explanations can appear fluent and pedagogically plausible, yet fail to describe the behavior of the GEC model. For applications such as writing assistance and second-language feedback, this mismatch matters because users may rely on the explanation to decide whether to accept a correction and how to revise similar errors in the future.

Existing explainable GEC work has introduced datasets and generation methods for evidence words, error types, and natural-language explanations \cite{fei-etal-2023-enhancing,song-etal-2024-gee,kaneko-okazaki-2024-controlled,ye2025excgec}. Other work has analyzed correction behavior at the edit level or used explanations as signals for retrieval \cite{ye-etal-2025-cleme2,li-etal-2025-explanation}. In parallel, broader explainability research has warned that explanation tests must distinguish plausibility, self-consistency, simulatability, and faithfulness \cite{hase-bansal-2020-evaluating,jacovi-goldberg-2020-towards,lyu-etal-2024-towards,parcalabescu-frank-2024-measuring}. The central question of this paper is therefore narrower: does an explanation predict the edit behavior of the GEC model that produced the correction?

We operationalize this question with a three-layer evaluation suite. L1 tests edit correspondence: given a source sentence and explanation, a reconstructor predicts the edit fields implied by the explanation. This layer is useful but leakage-prone, because an explicit explanation may simply copy the target edit. L2, Counterfactual Edit Simulatability, is the main behavioral test. We construct GEC-specific counterfactual variants, rerun the original correction model, and evaluate whether the explanation predicts whether the edit is preserved, changed, canceled, or retargeted. L3 tests rule and evidence grounding: whether the explanation states a grammar rule and evidence span consistent with the produced edit.

This framing deliberately avoids a stronger claim. Reconstruction and counterfactual simulatability do not reveal the internal causal mechanism of a neural GEC model. They are behavioral tests of whether an explanation is consistent with the model's observable edit decisions. An explanation can accurately describe a wrong correction and still be faithful to the model's behavior; conversely, a valid grammar rule may fail to explain the edit that was actually made. We therefore separate edit correspondence, behavioral faithfulness, grammatical validity, specificity, and learner helpfulness.

This paper contributes an executable pilot benchmark and diagnostic suite for this question, not a new explanation generator. We build a model-produced edit benchmark from EXPECT and JFLEG, run three public GEC model families, and derive explanation/control instances with explicit provenance. We implement L1 edit correspondence, L2 counterfactual edit simulatability, L3 rule/evidence grounding, leakage ablations, hard negatives, and an explanation-reranking application. The current evidence is deliberately bounded: all faithfulness labels in the benchmark are automatic, and no human-gold explanation labels are claimed.
