\section{RuleFaith-GEC}

RuleFaith-GEC generates and filters edit-level explanations through a sequence of explicit verification gates. Given a source sentence \(x\), a GEC model prediction \(\hat{y}\), and one model-produced atomic edit \(e\), the system produces a structured explanation candidate \(r\):
\[
r = G(x,\hat{y},e).
\]
Each candidate contains an edit description, an edit-validity judgment, a rule claim, evidence spans from the source sentence, applicability conditions, a rationale, confidence, and an optional abstention decision.

\input{../results/paper_assets/framework_figure}

\paragraph{Candidate generation.}
For each edit, a local teacher model produces multiple candidate styles, including natural, rule-grounded, evidence-grounded, concise, and uncertainty-aware explanations. The generator receives only the source, model prediction, atomic edit fields, and source-token span. It does not receive the reference correction, human labels, or the automatically inferred behavior category.

\paragraph{Edit validity gate.}
Before justifying an edit, RuleFaith asks whether the edit itself is valid, an acceptable alternative, invalid, stylistic, or uncertain. This gate is essential for wrong corrections and overcorrections. An explanation may be faithful to an invalid model edit, but it must not falsely present the invalid edit as grammatically required.

\paragraph{Edit alignment verifier.}
The alignment verifier checks whether the explanation corresponds to the current atomic edit rather than another edit in the same prediction. It evaluates span, source text, target text, operation, and direction. Reverse reconstruction is used here as an alignment diagnostic: if \(R(x,r)\) cannot recover the observed edit, the explanation is unlikely to identify the edit. However, reconstructability alone is not treated as evidence of rule correctness.

\paragraph{Rule and evidence verifiers.}
The rule verifier extracts or inspects the stated grammatical claim and tests whether it is plausible for the current edit type and source context. The evidence verifier checks whether cited evidence spans are source-located, sentence-specific, and sufficient for the rule. Source and target edit tokens are not automatically counted as contextual evidence; for instance, subject--verb agreement should cite the relevant subject, and article edits should cite the head noun or noun phrase context.

\paragraph{Leakage and specificity controls.}
RuleFaith separates required edit description from harmful answer copying. The edit-description field may state the observed operation, but the rule text and rationale are penalized when they merely copy the target or raw edit string. Target-masked validation removes the target string from rule, rationale, and condition fields and rechecks whether the remaining explanation still contains a grammar signal and source-specific evidence. Generic explanations that could fit any sentence are routed to refinement or rejection.

\paragraph{Targeted repair.}
When a candidate fails a dimension, RuleFaith creates a dimension-specific repair instruction. Alignment failures ask the generator to explain only the current edit. Validity failures require the explanation to distinguish an invalid or stylistic edit from a necessary correction. Rule failures request a rule that applies to the edit and avoids vague claims such as ``more natural''. Evidence failures request concrete source spans such as the subject, head noun, antecedent, governor, complement, or tense cue. The current pilot uses deterministic targeted repair for evidence integration and rationale edit-copy failures, then re-runs target-masked and rule/evidence gates.

\paragraph{Selection and abstention.}
Candidates are scored with separate alignment, validity, rule, evidence, specificity, leakage, and genericness terms. The system selects high-scoring candidates, sends repairable candidates through another verification pass, and rejects or abstains when no candidate has adequate rule and evidence support. The abstention output states that a reliable explanation cannot be generated for the displayed edit rather than supplying a fluent but unsupported rationale.

\paragraph{Relation to the stress-test metrics.}
The earlier L1/L2/L3 diagnostics remain part of the method infrastructure. L1 reconstruction evaluates edit correspondence and leakage. L2 counterfactual reruns test observable behavior under controlled source changes. L3 rule/evidence checks evaluate grammatical support. RuleFaith uses these diagnostics as gates for generation and repair, not as standalone claims that an explanation reveals the internal causal mechanism of the GEC model.
