\section{Related Work}

\paragraph{Grammatical error correction and edit analysis.}
GEC has been studied through shared tasks, neural generation models, edit-based models, and large language model prompting. Edit extraction and error typing are essential for fine-grained analysis. ERRANT introduced automatic edit extraction and rule-based error classification for GEC evaluation \cite{bryant-etal-2017-automatic}. CLEME2.0 further disentangles correction behavior into hit-correction, wrong-correction, under-correction, and over-correction for interpretable GEC evaluation \cite{ye-etal-2025-cleme2}. Our work uses this edit-level perspective but applies it to explanation faithfulness rather than correction scoring alone.

\paragraph{Explainable GEC and grammar error explanation.}
EXPECT annotates evidence words and error types to enhance GEC systems with explanations \cite{fei-etal-2023-enhancing}. GEE formulates grammar error explanation as generating one-sentence explanations for atomic edits and studies LLM-based pipelines across languages \cite{song-etal-2024-gee}. Prompt Insertion controls LLM explanation generation by inserting extracted correction points into the output process \cite{kaneko-okazaki-2024-controlled}. EXCGEC introduces an edit-wise explainable Chinese GEC benchmark with hybrid explanations \cite{ye2025excgec}. These studies motivate the need for explanations that correspond to edits. Our focus is an evaluation diagnostic: whether an explanation implies the same structured edit as the model output.

\paragraph{Explanations as signals for GEC.}
Explanation-based retrieval uses grammatical error explanations to select in-context demonstrations for multilingual GEC \cite{li-etal-2025-explanation}. This supports the broader view that explanations encode useful error-pattern information. Our work asks whether that information is faithful to a particular produced edit.

\paragraph{Faithfulness, simulatability, and counterfactuals.}
The broader explainability literature studies whether explanations reflect model behavior and whether users can simulate predictions from explanations \cite{hase-bansal-2020-evaluating,jacovi-goldberg-2020-towards,lyu-etal-2024-towards}. Recent work also warns that tests over natural-language explanations may measure output-level self-consistency rather than internal faithfulness \cite{parcalabescu-frank-2024-measuring}. Counterfactual and behavioral testing methods provide a way to probe model behavior under controlled changes \cite{ribeiro-etal-2020-beyond,gardner-etal-2020-evaluating,wu-etal-2021-polyjuice}. COCOGEC applies counterfactual generation to GEC robustness \cite{wang-etal-2026-cocogec}. Our distinction is to use counterfactual reruns to evaluate explanations of model-produced edits, not only correction robustness.
